Accesible, realistic genome simulation with selection using \stdpopsim

### 2025-03-06 analysis2_manuscript.tex
Update analysis2_manuscript.tex

    We evaluated the statistical power to detect selective sweeps for three methods:
    1) \sweepfinder \citep{degiorgio2016sweepfinder2},
    2) \diploshic \citep{kern2018diplos},
    and 3) reduced diversity ($\pi$).
    For each method, we determined a threshold for its test-statistic 
    using each of the two null models separately (neutral and BGS) and controlling 
    for $5\%$ false positive rate.
    We then applied each method to data simulated with sweeps for two sampled populations (YRI and CEU)
    in each of the 100 locations along human chromosome 1 (Figure \ref{fig:chr1_power}).
    First, we notice that power to detect sweeps is slightly lower when the test statistic threshold
    is determined using simulated data with background selection (BGS null model).
    %This observation is consistent with what was previously shown by \cite{schrider2020background},
    %but the effect is relatively minor. DRS: cutting this because my paper showed that power decreases when our test set has BGS and our training set does not, so not quite the same.
    This is likely because the strength of selection acting on exons is relatively weak, and the
    sweep detection methods we have used are relatively robust to the effects of background selection. 
    Second, the power to detect sweeps is generally higher in the YRI sample
    than in the CEU sample, which is consistent with the fact that YRI has a larger ancestral effective population size and
    thus both stronger selection and more variation to detect a sweep \citep[e.g.,][]{simonsen1995properties}.
    Further, the CEU population experienced the out-of-Africa bottleneck
    that may mute the signal of a sweep.  
    Finally and most importantly, the power to detect sweeps varies considerably along the chromosome,
    and this variation is much greater than the variation in power between CEU and YRI or between
    null models. This is likely due to the effects of recombination rate heterogeniety (Figure \ref{fig:chr1_power}).% as well
    %as the density of functional elements along the chromosome.
    %DRS: cut the density of func elements because that stuff doesn't seem to matter much here

@@

    One important finding we made using the new stdpopsim selection API is that inference of the DFE is sensitive to the dominance of deleterious mutations.
    All three of the inference methods of the DFE applied here assume that deleterious mutations act in an additive fashion (h=0.5) by default.
    However, the simulated DFE for the vaquita genome included highly recessive mutations.
    Specifically, these simulations used a DFE where mutations with $s>0.01$ are quite recessive, with $h<=0.01$ and where mutations with $0.001<s<0.01$ have $h=0.1$.
    We found that the three DFE inference methods underestimated $E[s]$ and overestimated the shape parameter (Fig 4).
    This performance is not surprising as decreasing $h$ for a given distribution of $s$ will lead to a decrease in $hs$,
    or an apparent decrease in the amount of selection.
    Fitting a mis-specified DFE that assumes $h=0.5$ will then lead to underestimates of $E[s]$.
    Indeed, the distribution of $hs$ is much more accurately estimated (Fig 4b).
    Overall, these results are consistent with prior work showing that it is challenging to separately infer the DFE of $s$ along with $h$ (cite PMID: 24830675; PMID: 39302992; PMID: 34951958 )
    and that a large proportion ($>5\%$) of recessive deleterious mutations can confound inferences of $s$ (PMID: 37074880).
    Given the plethora of evidence for recessive deleterious mutations (cite PMID: 4630587, PMID: 39114967 , PMID: 30013096 , PMID: 21098719 ),
    these results suggest the need for care in DFE inference.
    Kyriazis and Lohmueller (PMID: 39302992 ) inferred the DFE of $s$ for nonsynonymous mutations in humans when considering different values of $h$.
    Intriguingly, additive models show a satisfactory fit to the SFS of nonsynonymous variants in humans (Kyriazis et al) and more complex models of dominance do not substantially improve the fit.
    Nevertheless, models including recessive deleterious mutations, like those simulated here for the vaquita genome, show a similar fit to the SFS and may be more biologically realistic than fully additive models,
    given the evidence for the presence of recessive deleterious mutations.
    More broadly, new computational approaches are needed to precisely quantify the DFE and the covariance of $h$ and $s$.

@@

\section*{Funding}
    \label{funding}
    MFR, ST, NP, and ADK were supported in part by NIH award R35GM148253.
    RNG and LNT were supported in part by NIH award R35GM149235.
    DRS was supported in part by NIH award R35GM138286.

>>>>

    We evaluated the statistical power to detect selective sweeps for three methods:
    1) \sweepfinder \citep{degiorgio2016sweepfinder2},
    2) \diploshic \citep{kern2018diplos},
    and 3) reduced diversity ($\pi$).
    For each method, we determined a threshold for its test-statistic 
    using each of the two null models separately (neutral and BGS) and controlling 
    for a $5\%$ false positive rate.
    We then applied each method to data simulated with sweeps for two sampled populations (YRI and CEU)
    in each of the 100 locations along human chromosome 1 (Figure \ref{fig:chr1_power}).
    First, we notice that power to detect sweeps is slightly lower when the test statistic threshold
    is determined using simulated data with background selection (BGS null model).
    %This observation is consistent with what was previously shown by \cite{schrider2020background},
    %but the effect is relatively minor. DRS: cutting this because my paper showed that power decreases when our test set has BGS and our training set does not, so not quite the same.
    This is likely because the strength of selection acting on exons is relatively weak, and the
    sweep detection methods we have used are relatively robust to the effects of background selection. 
    Second, the power to detect sweeps is generally higher in the YRI sample
    than in the CEU sample, which is consistent with the fact that YRI has a larger ancestral effective population size and
    thus both stronger selection and more variation to detect a sweep \citep[e.g.,][]{simonsen1995properties}.
    Further, the CEU population experienced the out-of-Africa bottleneck
    that may mute the signal of a sweep.  
    Finally and most importantly, the power to detect sweeps varies considerably along the chromosome,
    and this variation is much greater than the variation in power between CEU and YRI or between
    null models. This is likely due to the effects of recombination rate heterogeniety (Figure \ref{fig:chr1_power}).% as well
    %as the density of functional elements along the chromosome.
    %DRS: cut the density of func elements because that stuff doesn't seem to matter much here

@@

    One important finding we made using the new stdpopsim selection API is that inference of the DFE is sensitive to the dominance of deleterious mutations.
    All three of the inference methods of the DFE applied here assume that deleterious mutations act in an additive fashion ($h=0.5$) by default.
    However, the simulated DFE for the vaquita genome included highly recessive mutations.
    Specifically, these simulations used a DFE where mutations with $s<-0.01$ are quite recessive, with $h<=0.01$ and where mutations with $-0.01<s<-0.001$ have $h=0.1$.
    We found that the three DFE inference methods underestimated $E[s]$ and overestimated the shape parameter (Fig 4).
    This performance is not surprising as decreasing $h$ for a given distribution of $s$ will lead to a decrease in $hs$,
    or an apparent decrease in the amount of selection.
    Fitting a mis-specified DFE that assumes $h=0.5$ will then lead to underestimates of $E[s]$.
    Indeed, the distribution of $hs$ is much more accurately estimated (Fig 4b).
    Overall, these results are consistent with prior work showing that it is challenging to separately infer the DFE of $s$ along with $h$ (cite PMID: 24830675; PMID: 39302992; PMID: 34951958 )
    and that a large proportion ($>5\%$) of recessive deleterious mutations can confound inferences of $s$ (PMID: 37074880).
    Given the plethora of evidence for recessive deleterious mutations (cite PMID: 4630587, PMID: 39114967 , PMID: 30013096 , PMID: 21098719 ),
    these results suggest the need for care in DFE inference.
    Kyriazis and Lohmueller (PMID: 39302992 ) inferred the DFE of $s$ for nonsynonymous mutations in humans when considering different values of $h$.
    Intriguingly, additive models show a satisfactory fit to the SFS of nonsynonymous variants in humans (Kyriazis et al) and more complex models of dominance do not substantially improve the fit.
    Nevertheless, models including recessive deleterious mutations, like those simulated here for the vaquita genome, show a similar fit to the SFS and may be more biologically realistic than fully additive models,
    given the evidence for the presence of recessive deleterious mutations.
    More broadly, new computational approaches are needed to precisely quantify the DFE and the covariance of $h$ and $s$.

@@

\section*{Funding}
    \label{funding}
    MFR, ST, NP, and ADK were supported in part by NIH award R35GM148253.
    RNG and LNT were supported in part by NIH award R35GM149235.
    DRS was supported in part by NIH award R35GM138286.
    KEL was supported by NIH award R35GM119856.
